Surrogate model for an aligned-spin effective one body waveform model of binary neutron star inspirals using Gaussian process regression

### 2018-12-02 paper/paper_no_event.tex
Paper: edits in Sec V

The evaluation time has a flat cost of $\sim 0.01$~s to perform the GPR interpolation at each node. The rest of the time is spent resampling the waveform with spline interpolation. For a starting frequency of 20~Hz, this takes 0.13~s when the waveform is matched with data uniformly sampled at 4096Hz. For current parameter estimation codes, this is sufficient. However, one could further improve run times using reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} which requires frequency-domain waveforms, or multi-band waveform interpolation~\cite{VinciguerraVeitchMandel2017}. Finally, we note that the production MCMC sampler used for the GW170817 analysis has an autocorrelation length of $\mathcal{O}(10^3)$ for aligned-spin BNS models, so significant improvements to the parameter estimation runtime can be made through better samplers.

The hierarchical method presented here, where we begin with the analytic TaylorF2 reference model then make a surrogate of the residual, can be used to further improve the waveform model. For example, one could build a surrogate for numerical BNS simulations using \texttt{SEOBNRv4T\_surrogate} as the base model, and constructing a surrogate of the residual. Such a model would have the accuracy of EOB below $\sim 400$~Hz and the accuracy of numerical simulations for the last several cycles before merger. Current state-of-the-art NR simulations have phase errors of several tenths of a radian over the last $\sim 20$ gravitational-wave cycles~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}. Using GPR and uncertainty sampling discussed above, one could optimally choose the waveform parameters for the numerical simulations, and run them in parallel to build the training set. Because the difference between \texttt{SEOBNRv4T} and numerical BNS simulations is exceptionally small, one would not need a high fractional accuracy for a surrogate of the difference, and 10--100 waveforms may be sufficient.

This model notably does not include precession for non-aligned spins. Although none of the EOB models currently available include both tidal effects and precession, there are ways to boost this waveform model to a precessing frame and approximately incorporate precession. For example, Chatziioanno {\it et al}.\ have analytically solved the 2PN-accurate precession equations for generic spins~\cite{ChatziioannouKleinCornish2017a, ChatziioannouKleinCornish2017b}, with the exception of transitional precession which is unlikely for low mass-ratio and spin BNS systems. Using stationary uniform asymptotics, they can also analytically Fourier-transform the solution. Importantly, one is free to specify the frequency-domain amplitude and phase evolution, such as the \texttt{SEOBNRv4T\_surrogate} here, for the waveform in the co-precessing frame. This approach would provide a fast, accurate model for BNS systems with tides and generic spins. We note, however, that including precession did not noticeably affect the parameter estimation results for GW170817~\cite{BNSPE}.

>>>>

The evaluation time has a flat cost of $\sim 0.01$~s to perform the GPR interpolation at each node. The rest of the time is spent resampling the waveform with spline interpolation. For a starting frequency of 20~Hz, this takes 0.13~s when the waveform is matched with data uniformly sampled at 4096Hz. For current parameter estimation codes, this is sufficient. However, one could further improve run times using reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} which requires frequency-domain waveforms, multi-band waveform interpolation~\cite{VinciguerraVeitchMandel2017} or relative binning~\cite{Zackay:2018qdy}. Finally, we note that the production MCMC sampler used for the GW170817 analysis has an autocorrelation length of $\mathcal{O}(10^3)$ for aligned-spin BNS models, so significant improvements to the parameter estimation runtime can be made through better samplers.

The hierarchical method presented here, where we begin with the analytic TaylorF2 reference model then make a surrogate of the residual, can be used to further improve the waveform model. For example, one could build a surrogate for numerical BNS simulations using \texttt{SEOBNRv4T\_surrogate} as the base model, and constructing a surrogate of the residual. Such a model would have the accuracy of EOB below $\sim 400$~Hz and the accuracy of numerical simulations for the last several cycles before merger. Current state-of-the-art NR simulations have phase errors of several tenths of a radian over the last $\sim 20$ gravitational-wave cycles~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}. Using GPR and uncertainty sampling discussed above, one could optimally choose the waveform parameters for the numerical simulations, and run them in parallel to build the training set. If the difference between \texttt{SEOBNRv4T} and numerical BNS simulations is small, one would not need a high fractional accuracy for a surrogate of the difference, and 10--100 waveforms may be sufficient.

This model notably does not include precession for non-aligned spins. Although none of the EOB models currently available include both tidal effects and precession, there are ways to rotate this waveform model to a precessing frame and approximately incorporate precession. For example, Chatziioannou {\it et al}.\ have analytically solved the 2PN-accurate precession equations for generic spins~\cite{ChatziioannouKleinCornish2017a, ChatziioannouKleinCornish2017b}, with the exception of transitional precession which is unlikely for low mass-ratio and spin BNS systems. Using shifted uniform asymptotics~\cite{Klein:2014bua}, they can also analytically Fourier-transform the solution. Importantly, one is free to specify the frequency-domain amplitude and phase evolution, such as the \texttt{SEOBNRv4T\_surrogate} here, for the waveform in the co-precessing frame. This approach would provide a fast, accurate model for BNS systems with tides and generic spins. We note, however, that including precession did not noticeably affect the parameter estimation results for GW170817~\cite{BNSPE}.
